### 2021-04-14 conclusion.tex
Added remarks about GV->PGV.

\subsubsection*{Ties with logic}
The correspondence between logic and types lays the foundation for functional programming~\cite{wadler15}.
Since its inception by Girard~\cite{girard87}, linear logic has been a candidate for a foundational correspondence for concurrent programs.
A~correspondence with linear $\pi$-calculus was established early on by Abramsky~\cite{abramsky94} and Bellin and Scott~\cite{bellinscott94}. Many years later, correspondences with the $\pi$-calculus with binary session types were proposed. Caires and Pfenning~\cite{cairespfenning10} propose a correspondence with dual intuitionistic linear logic, while Wadler~\cite{wadler14} proposes a correspondence with classical linear logic. Both guarantee deadlock freedom via their connection to logic, as a consequence of cut elimination.
Dardha and Gay~\cite{dardhagay18extended} integrate Kobayashi and Padovani's work on priorities~\cite{kobayashi06,padovani14} with CP, loosening its ties to linear logic in exchange for expressivity.
Dardha and P\'{e}rez~\cite{dardhaperez15extended} compare Kobayashi-style typing and CLL typing, and show that CLL corresponds to a subsystem of Kobayashi's where restriction is applied once.
Balzer~\etal~\cite{balzerpfenning17} introduce sharing, which breaks deadlock freedom.
They later restore deadlock freedom using priorities~\cite{balzertoninho19}.
Carbone~\etal~\cite{CarboneMSY15,carbonelindley16} give a logical view of MPST with a generalised duality.
Caires and P\'{e}rez~\cite{CairesP16} give a presentation of MPST in terms of binary session types and the use of a \emph{medium process} which guarantee protocol fidelity and deadlock freedom.
Their binary session types are rooted in linear logic following the propositions-as-sessions correspondences.
Ciobanu and Horne~\cite{CiobanuH15} gave the first propositions-as-sessions correspondence of MPST with the \emph{calculus of structures}~\cite{Guglielmi07} with global and local session types are inspired by Scribble, which determines whether local session types can be combined in a deadlock free manner.
Horne~\cite{Horne20} presents a system for subtyping and multiparty compatibility where compatible processes are race free and deadlock free using a propositions-as-sessions correspondence, similar to the approach in~\cite{CiobanuH15}, with a non-commutative extension of linear logic.

>>>>

\subsubsection*{Ties with logic}
The correspondence between logic and types lays the foundation for functional programming~\cite{wadler15}.
Since its inception by Girard~\cite{girard87}, linear logic has been a candidate for a foundational correspondence for concurrent programs.
A~correspondence with linear $\pi$-calculus was established early on by Abramsky~\cite{abramsky94} and Bellin and Scott~\cite{bellinscott94}. Many years later, several correspondences between linear logic and the $\pi$-calculus with binary session types were proposed. Caires and Pfenning~\cite{cairespfenning10} propose a correspondence with dual intuitionistic linear logic, while Wadler~\cite{wadler14} proposes a correspondence with classical linear logic. Both guarantee deadlock freedom as a consequence of cut elimination.
Dardha and Gay~\cite{dardhagay18extended} integrate Kobayashi and Padovani's work on priorities~\cite{kobayashi06,padovani14} with CP, loosening its ties to linear logic in exchange for expressivity.
Dardha and P\'{e}rez~\cite{dardhaperez15extended} compare priorities \`a la Kobayashi with tree restrictions \`a la CP, and show that the latter is a subsystem of the former.
Balzer~\etal~\cite{balzerpfenning17} introduce sharing at the cost of deadlock freedom, which they restore using an approach similar to priorities~\cite{balzertoninho19}.
Carbone~\etal~\cite{CarboneMSY15,carbonelindley16} give a logical view of MPST with a generalised duality.
Caires and P\'{e}rez~\cite{CairesP16} give a presentation of MPST in terms of binary session types and the use of a \emph{medium process} which guarantee protocol fidelity and deadlock freedom. Their binary session types are rooted in linear logic.
Ciobanu and Horne~\cite{CiobanuH15} give the first Curry-Howard correspondence between MPST and BV~\cite{Guglielmi07}, a conservative extension of linear logic with a non-commutative operator for sequencing.
Horne~\cite{Horne20} give a system for subtyping and multiparty compatibility where compatible processes are race free and deadlock free using a Curry-Howard correspondence, similar to the approach in~\cite{CiobanuH15}.

### 2021-04-14 introduction.tex
Minor changes.
Added remarks about GV->PGV.

In the $\pi$-calculus literature, a growing line of work leverages the Curry-Howard correspondences between session-typed $\pi$-calculus and linear logic~\cite{girard87}: Caires and Pfenning's $\pi$DILL~\cite{cairespfenning10} corresponds to dual intuitionistic linear logic~\cite{barber96}, and Wadler's Classical Processes~\cite[CP]{wadler14} corresponds to classical linear logic~\cite[CLL]{girard87}. Both calculi guarantee deadlock freedom, which they achieve by restricting structure of processes and shared channels to \emph{trees}, by combing name restriction and parallel composition into a single construct, corresponding to the logical cut. This ensures that two processes can only communicate via exactly one series of channels, which rules out interleavings of sessions, and guarantees deadlock freedom.
There are many downsides to combining name restriction and parallel composition---lack of modularity, difficulty typing structural congruence and formulating label-transition semantics, \etc---which have led to various approaches to decoupling these constructs. Hypersequent CP~\cite{MP18,kokkemontesi19popl,kokkemontesi19tlla} and Linear Compositional Choreographies~\cite{CarboneMS18} decouple them, but maintain the correspondence to CLL and allow only tree-structured processes. Priority CP~\cite[PCP]{dardhagay18extended} weakens the correspondence to CLL in exchange for a more expressive language which allows cyclic-structured processes. PCP decouples CP's cut rule into two separate constructs: one for parallel composition via a mix rule, and one for name restriction via a cycle rule. To restore deadlock freedom, PCP uses \emph{priorities}~\cite{kobayashi06,padovani14}. Priorities encode the \emph{order of actions} and rule out bad interleavings. Dardha and Gay~\cite{dardhagay18extended} prove cycle-elimination for PCP, adapting the cut-elimination proof for classical linear logic, and deadlock freedom follows as a corollary.

@@

Our aims are to develop the expressiveness of GV but maintain deadlock freedom. We chose GV for several reasons, some of which apply more generally as benefits to working within a concurrent $\lambda$-calculus as opposed to a name-passing process calculus. Concurrent $\lambda$-calculi support higher-order functions, and have a capability for abstraction not usually present in process calculi. Within a concurrent $\lambda$-calculus, one can derive extensions of the communication capabilities of the language via well-understood extensions of the functional fragment, \eg, we can derive internal/external choice from sum types. Concurrent $\lambda$-calculi maintain a clear separation between the program which the user writes and the configurations which represent the state of the system as it evaluates the program. However, our main motivation is that results obtained for $\lambda$-calculi transfer more easily to real-world functional programming languages. \emph{Case in point}: we were able to embed the theory developed in this paper in Linear Haskell~\cite{bernardyboespflug18} without much effort~\cite{KD21b}. For us, these advantages outweigh the benefits of process calculi, \eg, their succinctness.

The benefits of working with GV specifically, as opposed to other concurrent $\lambda$-calculi, is its relation to CP~\cite{wadler14}, and its strong formal properties. We thus pose our research question for GV:
\begin{quotation}
  \textbf{RQ:}
  Can we design a more expressive GV which guarantees deadlock freedom for cyclic-structured processes?
\end{quotation}

>>>>

In the $\pi$-calculus literature, there is a growing line of work in developing Curry-Howard correspondences between session-typed $\pi$-calculus and linear logic~\cite{girard87}: Caires and Pfenning's $\pi$DILL~\cite{cairespfenning10} corresponds to dual intuitionistic linear logic~\cite{barber96}, and Wadler's Classical Processes~\cite[CP]{wadler14} corresponds to classical linear logic~\cite[CLL]{girard87}. Both calculi guarantee deadlock freedom, which they achieve by restricting structure of processes and shared channels to \emph{trees}, by combing name restriction and parallel composition into a single construct, corresponding to the logical cut. This ensures that two processes can only communicate via exactly one series of channels, which rules out interleavings of sessions, and guarantees deadlock freedom.
There are many downsides to combining name restriction and parallel composition---lack of modularity, difficulty typing structural congruence and formulating label-transition semantics, \etc---which have led to various approaches to decoupling these constructs. Hypersequent CP~\cite{MP18,kokkemontesi19popl,kokkemontesi19tlla} and Linear Compositional Choreographies~\cite{CarboneMS18} decouple them, but maintain the correspondence to CLL and allow only tree-structured processes. Priority CP~\cite[PCP]{dardhagay18extended} weakens the correspondence to CLL in exchange for a more expressive language which allows cyclic-structured processes. PCP decouples CP's cut rule into two separate constructs: one for parallel composition via a mix rule, and one for name restriction via a cycle rule. To restore deadlock freedom, PCP uses \emph{priorities}~\cite{kobayashi06,padovani14}. Priorities encode the \emph{order of actions} and rule out bad interleavings. Dardha and Gay~\cite{dardhagay18extended} prove cycle-elimination for PCP, adapting the cut-elimination proof for classical linear logic, and deadlock freedom follows as a corollary.

@@

Our aims are to develop the expressiveness of GV but maintain deadlock freedom. We chose GV for several reasons, some of which apply more generally as benefits to working within a concurrent $\lambda$-calculus as opposed to a name-passing process calculus. Concurrent $\lambda$-calculi support higher-order functions, and have a capability for abstraction not usually present in process calculi. Within a concurrent $\lambda$-calculus, one can derive extensions of the communication capabilities of the language via well-understood extensions of the functional fragment, \eg, we can derive internal/external choice from sum types. Concurrent $\lambda$-calculi maintain a clear separation between the program which the user writes and the configurations which represent the state of the system as it evaluates the program. However, our main motivation is that results obtained for $\lambda$-calculi transfer more easily to real-world functional programming languages. \emph{Case in point}: we were able to embed the theory developed in this paper in Linear Haskell~\cite{bernardyboespflug18} without much effort~\cite{KD21b}. For us, these advantages outweigh the benefits of process calculi, \eg, their succinctness.
The benefits of working with GV specifically, as opposed to other concurrent $\lambda$-calculi, is its relation to CP~\cite{wadler14}, and its strong formal properties. We thus pose our research question for GV:
\begin{quotation}
  \textbf{RQ:}
  Can we design a more expressive GV which guarantees deadlock freedom for cyclic-structured processes?
\end{quotation}

### 2021-04-14 priority-gv.tex
Added remarks about GV->PGV.

We present Priority GV (PGV), a session-typed functional language based on GV~\cite{wadler15,lindleymorris15} which uses priorities \`{a} la Kobayashi and Padovani~\cite{kobayashi06,padovaninovara15} to enforce deadlock freedom.
Priority GV offers a more fine-grained analysis of communication structures, and by separating channel creation form thread spawning it allows cyclic structures.
We illustrate this with two programs in PGV, \cref{ex:cycle,ex:deadlock}. Each program contains two processes---the main process, and the child process created by $\tm{\spawn}$---which communicate using \emph{two} channels. The child process receives a unit over the channel $\tm{x}/\tm{x'}$, and then sends a unit over the channel $\tm{y}/\tm{y'}$. The main process does one of two things:
\begin{enumerate*}[label= (\alph*) ]
\item in \cref{ex:cycle}, it sends a unit over the channel $\tm{x}/\tm{x'}$, and then waits to receive a unit over the channel $\tm{y}/\tm{y'}$;
\item in \cref{ex:deadlock}, it does these in the opposite order, which results in a deadlock.
\end{enumerate*}
We will show that only \cref{ex:cycle} is typeable in PGV, while it is not typable in GV~\cite{wadler14} nor guaranteed to be deadlock-free by its predecessor~\cite{gayvasconcelos12}.

>>>>

We present Priority GV (PGV), a session-typed functional language based on GV~\cite{wadler15,lindleymorris15} which uses priorities \`{a} la Kobayashi and Padovani~\cite{kobayashi06,padovaninovara15} to enforce deadlock freedom.
Priority GV offers a more fine-grained analysis of communication structures, and by separating channel creation form thread spawning it allows cyclic structures.
We illustrate this with two programs in PGV, \cref{ex:cycle,ex:deadlock}. Each program contains two processes---the main process, and the child process created by $\tm{\spawn}$---which communicate using \emph{two} channels. The child process receives a unit over the channel $\tm{x}/\tm{x'}$, and then sends a unit over the channel $\tm{y}/\tm{y'}$. The main process does one of two things:
\begin{enumerate*}[label= (\alph*) ]
\item in \cref{ex:cycle}, it sends a unit over the channel $\tm{x}/\tm{x'}$, and then waits to receive a unit over the channel $\tm{y}/\tm{y'}$;
\item in \cref{ex:deadlock}, it does these in the opposite order, which results in a deadlock.
\end{enumerate*}
PGV is more expressive than GV: \cref{ex:cycle} is typeable and guaranteed deadlock-free in PGV, but is not typeable in GV~\cite{wadler14} and not guaranteed deadlock-free in GV's predecessor~\cite{gayvasconcelos12}. We believe PGV is a non-conservative \emph{extension} of GV, as CP can be embedded in a Kobayashi-style system~\cite{dardhaperez15extended}.
